---
title: "Covid91 vaccine study Final2024"
author: "Onalee Garusinghe"
date: "11/26/24"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
  pdf_document:
    toc: true
    toc_depth: '5'
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, covid19 = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
males = filter(FinalData, sex =="M")
females = filter(FinalData, sex =="F")
gay = filter(FinalData, LGBTQ =="gay")
Druguser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

##OverAll Approach

1. **Descriptive Analysis**: Summarize data visually then numerically.
2. **Inferential Analysis**: Conduct hypothesis testing:
   - \(H_0\): The vaccine has no impact on infection rates.
   - \(H_A\): The vaccine reduces infection rates.
3. **Effectiveness Calculation**:
   \[
   \text{Vaccine Effectiveness (VE)} = 1 - \frac{\text{Risk}_{\text{vaccine}}}{\text{Risk}_{\text{placebo}}}
   \]
4. **Effectivness Confidence Interval**: low to hi

## Males

```{r}
dd2 <- males %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

   This plot represents the population of males and their relationship to getting covid-19 by taking either the vaccine or the placebo. The plot shows that around 68% of males who got covid took the placebo. Whereas about 25% of males who got covid took the vaccine. Therefore, more males who got covid took the placebo, whereas the males who got covid from taking the vaccine were not as likely to contract covid as much. This means that the vaccine seems to be effective for the males in this dataset.There were two plots that came out. The first plot shows two bars for the treatment, either the placebo or the vaccine. Along the x-axis was the amount of people there was, and along the y-axis shows the infected from covid, either placebo or vaccine. It shows the number of people infected from covid or from the placebo.

## Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=males)
rowPerc(table1)
colPerc(table1)
```

  The numerical results for the males shows the percentages of males who got covid while taking the vaccine and males who got covid while taking the placebo. It shows that around 74% of the males who got covid took the placebo. Whereas only 26% of males who got covid took the vaccine. This comes to show that the vaccine is effective towards males. 

### Inferential 

```{r}
chisq.test(table1)
chisqtestGC(table1)
malesfish <- fisher.test(table1)
malesfish
```

A Fisher’s Exact test was run and a p-value was given. The p-value came out to be 2.2-16, which is quite a small p-value. This means that these results are not likely to occur by chance and that there is a statistically significant result.
The odds ratio above came out to be 2.81, meaning that the males that took the vaccine were less likely to contract covid than those who took the placebo.

```{r}
effect <- malesfish$p.value/(malesfish$p.value +1)
effectivenessguy = malesfish$estimate/(malesfish$estimate +1)*100
print(effectivenessguy)
```

```{r}
low=(malesfish$conf.int[1]/(malesfish$conf.int[1] +1))*100
hi=(malesfish$conf.int[2]/(malesfish$conf.int[2] +1))*100
options(digits=2)
```

## Effectiveness Confidence Interval `r low`  to `r hi`

The odds ratio essentially shows how effective the vaccine is. The odds ratio for the males is around 73.8, with an effectiveness confidence interval of low being  68.84-78.32% hi, meaning that the vaccine is effective in protecting males from getting covid-19.

## Females

```{r}
dd2 <- females %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

  This plot represents the population of females and their relationship to getting covid-19 by taking either the vaccine or the placebo. The plot shows that around 58% of females who got covid took the placebo. Whereas about 37% of females who got covid took the vaccine.  Therefore, more females who got covid took the placebo. This means that the vaccine seems to be effective for females in this dataset. Although, it seems to be more effective for males than females, since more females than males who contracted covid took the vaccine rather than the placebo.There were two plots that came out. The first plot shows two bars for the treatment, either the placebo or the vaccine. Along the x-axis was the amount of people there was, and along the y-axis shows the infected from covid, either placebo or vaccine. It shows the number of people infected from covid or from the placebo. The second plot also shows two bars for the treatments, either the placebo or the vaccine. It shows the percentage of people infected from covid or from the placebo.

## Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=females)
rowPerc(table1)
colPerc(table1)
```

The numerical results for the females shows the percentages of females who got covid while taking the vaccine and males who got covid while taking the placebo. It shows that around 65% of the females who got covid took the placebo. Whereas only 35% of females who got covid took the vaccine. This comes to show that the vaccine is effective towards females.

## Inferential 

```{r}
chisq.test(table1)
chisqtestGC(table1)
femalesfish <- fisher.test(table1)
femalesfish
```

  A Fisher’s Exact test was run and a p-value was given. The p-value came out to be 1e-08, which is also quite a small p-value. This means that these results are not likely to occur by chance and that there is a statistically significant result.	The odds ratio above came out to be 1.9, meaning that the females that got covid by taking the vaccine were less likely to contract covid than those who took the placebo. The odds of getting covid while taking the vaccine is 2:1.
  
```{r}
effect <- femalesfish$p.value/(femalesfish$p.value +1)
effectivenessfemales = femalesfish$estimate/(femalesfish$estimate +1)*100
print(effectivenessfemales)
```

```{r}
low=(femalesfish$conf.int[1]/(femalesfish$conf.int[1] +1))*100
hi=(femalesfish$conf.int[2]/(femalesfish$conf.int[2] +1))*100
options(digits=2)
```

## Effectiveness Confidence Interval `r low`  to `r hi`

Shown above is the effectiveness for the covid-19 vaccine for females. If the effectiveness is above 60%, then the vaccine is shown to be effective. Based on the effectiveness confidence interval of low as 60%-70% hi, I can conclude that the vaccine is effective in protecting females from getting covid-19.The odds ratio essentially shows how effective the vaccine is. The odds ratio for the females is 65 with an effectiveness confidence interval of 59.8 to 69.9%, meaning that the vaccine is effective in protecting females from getting covid-19.

## LGBTQ

```{r}
dd2 <- gay %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

  This plot represents the population of gays and their relationship to getting covid-19 by taking either the vaccine or the placebo. The plot shows that around 80% of gays who got covid took the placebo. Whereas about 20% of gays who got covid  took the vaccine. Therefore, more gays who got covid took the vaccine, whereas the gays who got covid by taking the placebo did not contract covid as much. This means that the vaccine does not seem to be effective for the gays in this data set. 
There were two plots that came out. The first plot shows two bars for the treatment, either the placebo or the vaccine. Along the x-axis was the amount of people there was, and along the y-axis shows the infected from covid, either placebo or vaccine. It shows the number of people infected from covid or from the placebo. The second plot also shows two bars for the treatments, either the placebo or the vaccine. It shows the percentage of people infected from covid or from the placebo.

## Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=gay)
rowPerc(table1)
colPerc(table1)
```

  The numerical results for the gays shows the percentages of gays who got covid while taking the vaccine and gays who got covid while taking the placebo. It shows that 80% of the gays who got covid took the placebo. Whereas 20% of the gays who got covid took the vaccine. More gays who took the vaccine got covid. This comes to show that the vaccine is ineffective towards the gays at preventing the contraction of covid-19.
  
## Inferential

```{r}
chisq.test(table1)
chisqtestGC(table1)
gayfish <- fisher.test(table1)
gayfish
```

 A Fisher’s Exact test was run and a p-value was given. The p-value came out to be 7e-07, which is a small p-value. This means that these results are not likely to occur by chance and that there is a statistically significant result.
The null hypothesis is rejected in favor of the alternative hypothesis, meaning that the vaccine is ineffective in preventing the likelihood of getting covid in individuals who identify as gay.
	The odds ratio above came out to be 4.3, meaning that those gays that got covid while taking the vaccine were more likely to contract covid than those who took the placebo and got covid. The odds of getting covid-19 while taking the vaccine is 11.3, meaning there is a big chance at contracting covid while taking the vaccine for the gays.
	
```{r}
effect <- gayfish$p.value/(gayfish$p.value +1)
effectivenessgay = gayfish$estimate/(gayfish$estimate +1)*100
print(effectivenessgay)
```

```{r}
low=(gayfish$conf.int[1]/(gayfish$conf.int[1] +1))*100
hi=(gayfish$conf.int[2]/(gayfish$conf.int[2] +1))*100
options(digits=2)
```

## Effectiveness Confidence Interval `r low`  to `r hi`

Shown above is the effectiveness for the covid-19 vaccine for the gays. If the effectiveness is exactly 65%, then the vaccine is shown to be effective. Based on the effectiveness confidence interval as low as 68.99%-89.87% hi, I can conclude that the vaccine is ineffective in protecting the gays from getting covid-19.
	The odds ratio essentially shows how effective the vaccine is. The odds ratio for the gays is 81, with an effectiveness confidence interval of 68.9-89.8%, meaning that the vaccine is not effective in protecting the gays from getting covid-19.
	
## Druggies

```{r}
dd2 <- Druguser %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

This plot represents the population of drug users and their relationship to getting covid-19 by taking either the vaccine or the placebo. The plot shows that 5% of the drug users who got covid took the placebo. Whereas 83% of the drug users who got covid took the vaccine. Therefore, more gays who got covid took the vaccine, whereas the gays who got covid while taking the placebo did not contract covid as much. This means that the vaccine does not seem to be effective for the drug users in this dataset. 
	There were two plots that came out. The first plot shows two bars for the treatment, either the placebo or the vaccine. Along the x-axis was the amount of people there was, and along the y-axis shows the infected from covid, either placebo or vaccine. It shows the number of people infected from covid or from the placebo. The second plot also shows two bars for the treatments, either the placebo or the vaccine. It shows the percentage of people infected from covid or from the placebo.


	
# Overall Results and Conclusions